Global Optimization of an Analog Method by Means of Genetic Algorithms

### 2016-09-06 Genetic algorithms - MWR/template.tex
GAs paper under correction.

Analog methods (AMs) relies on the hypothesis that similar situations in terms of atmospheric circulation are likely to lead to similar local weather \citep{Lorenz1956, Lorenz1969, Duband1970, Bontron2005}. The principle consists in sampling a certain number of past situations based on different atmospheric variables (predictors) in order to build a probabilistic prediction for a local weather variable of interest (predictand). A common usage of AMs is for precipitation forecasting or downscaling \citep[eg.][]{Guilbaud1997, Bontron2005, Hamill2006, Bliefernicht2010, Marty2012, Horton2012, Radanovics2013, Chardon2014, Dayon2015, Hamill2015b, BenDaoud2016}, but AMs, or equivalent, were also used to predict temperature \citep{Radinovic1975, Woodcock1980, Kruizinga1983, DelleMonache2013, Caillouet2016}, wind \citep{Gordon1987, DelleMonache2013, DelleMonache2011, Vanvyve2015, Alessandrini2015, Junk2015, Junk2015c}, solar power \citep{Alessandrini2015a, Bessa2015}, snow avalanches \citep{Obled1980, Bolognesi1993}, insolation \citep{Bois1981}, and the trajectory of tropical cyclones \citep{Keenan1981, Sievers2000, Fraedrich2003}. Applications for monthly time series in many countries also exist, including Canada \citep{Shabbar1986},  Hungary \citep{Toth1989}, the Netherlands \citep{Nap1981}, and England \citep{Murray1974}, as well as seasonal predictions: \citet{Barnett1978}, \citet{Bergen1982} and \citet{Livezey1988}.

Even though the method is rather simple, it contains a certain number of parameters one needs to determine, such as the choice of the predictor variable, its pressure level and temporal window (hour of observation of the predictor) to consider, the spatial domain to use for the comparison, as well as the analogy criteria itself, and finally the number of analog situations to keep at each subsampling level.

@@

In order to overcome these limitations, two optimization techniques were assessed. First, \citet{Horton2012a} assessed the ability of the \citet{Nelder1965a} method based on a simplex algorithm. This technique did not provide satisfying results and failed at converging toward a unique solution. The parameter space of AMs can be very complex and is inappropriate for a linear optimization technique. The conclusion was that global optimization techniques were necessary in order to calibrate most AM variants, as it is the only way to optimize all parameters of all analogy levels simultaneously. In addition, it can overcome the systematic manual assessments of all pressure levels and temporal windows. Finally, it can open new perspectives by allowing the addition of new degrees of freedom, such as a weighting of the criteria values between the pressure levels, and the consideration of differentiated spatial windows between the pressure levels. The relevance of Genetic Algorithms (GAs) is presented here, which does not exclude that other global optimization techniques could eventually be successful. If using GAs to optimize AMs is computationally intensive, once an AM is calibrated, its use in real-time operations it very fast and lightweight.

@@

Predictors, for precipitation prediction, can be varied: e.g. geopotential heights at different pressure levels and on different temporal windows (time of observation), humidity variables, etc \citep[see][for a more detailed list]{BenDaoud2016}. The method is often made of several levels of analogy, leading to successive subsampling on predictors of different nature (eg. atmospheric circulation, moisture variables, vertical motion, and air temperature).

@@

	\item First level of analogy: $n_{1}$ dates are selected out of the preselected $n_{0}$, by means of an analogy ranking. The first level of analogy is often based on the atmospheric circulation when it comes to precipitation prediction. The similarity of the atmospheric circulation of the target date with the candidate situations is thus assessed on the geopotential height (at specific pressure levels, such as 500 and 1000~hPa, and over a certain spatial window, e.g. 12~h and 24~h) by means of the S1 criteria \citep[Eq. (\ref{eq:S1}), ][]{Teweles1954, Drosdowsky2003}, which is a comparison of gradients, over a certain spatial gridded window. This criteria was found more relevant than Euclidean distances for pressure fields by different studies \citep{Wilson1980, Woodcock1980, Guilbaud1998, Bontron2004}.

@@

\begin{equation}
\label{eq:CRPSS}
CRPSS = \frac{CRPS-CRPS_{r}}{CRPS_{p}-CRPS_{r}} = 1-\frac{CRPS}{CRPS_{r}}
\end{equation}
where $CRPS_{r}$ is the CRPS value for the reference and $CRPS_{p}$ would be the one for a perfect prediction (which implies $CRPS_{p}~=~0$). A better prediction is characterized by an increase in CRPSS.

@@

A key element of GAs configuration is finding the right balance between exploration and exploitation \citep{Back1992a, Smith1997a}. Exploration is characterized by a relatively high probability to assess the regions of the parameter space that have not yet been visited. This probability must be sufficiently large at the beginning of the optimization, so that the algorithm is capable of identifying the region where the global optimum is likely located. Exploitation is characterized by a local search in an area of interest, and generally makes small movements. The latter is interesting to refine the results at the end of the optimization.

@@

GAs optimize a population of $N$ individuals (parameter sets). Each individual contains a chromosome (set of parameters of our AM in this case). Genes are the individual parameters constituting the chromosome. They can be either categorical (e.g. choice of the meteorological variable), discrete (e.g. number of analogs to select), or continuous.

@@

The mutation operator is a direct modification of some gene values. It adds diversity to the population and prevents a freeze of the evolution, or a genetic drift to a local optimum. Thus, it makes the convergence to the global optimum theoretically possible \citep{Beasley1993a}, as it allows exploring beyond the current region of the parameters space by bringing new characteristics that were not present in the original population \citep{Haupt2004}.

@@

As the assessment (calculation of the objective function) of each member of the population of a generation is completely independent, it was performed in parallel on different processors of a computer. In order to perform optimizations for multiple time series, the use of a cluster is a necessity, which our code allows.

Some constraints need to be taken into account. For example, when a crossover or a mutation operation results in a parameter value standing out of the authorized bounds, it has to be brought back within the limits. Moreover, the parameters are of different nature: some are continuous, some are discrete, and finally, some are categorical, i.e. independent elements in an array, such as the selection of the meteorological variable. New values resulting from the optimizer need to respect the type of data it represents. Other constraints exist in between the parameters, e.g. the hours of observation of the relative humidity and the precipitable water predictors have to match when processing the moisture index.

@@

One of our goals being to make recommendations of how to use GAs in view of optimizing AM variants, a systematic procedure was adopted. The results are summarized hereafter \citep[see][for the details]{Horton2012a}. In order to compare different configurations of GAs, the factorial design approach was applied, in a similar way as \citet{Costa2005a,Costa2007a} and \citet{Mariano2010a}. It allows isolating the effect of a parameter under different combinations of the other options. A procedure by stages was adopted, analyzing in details and in a systematic way every variant of the implemented operators, in combination with multiple other options and parameters in order to take into account eventual co-dependencies.

@@

The performances were characterized by four indicators: (i) mean performance score: average of the final scores of the 10 optimizations, (ii) convergence: the number of optimizations that converged to a supposed global optimum, (iii) number of generations: characterization of the convergence speed, and (iv) number of evaluations of the objective function: characterization of the required calculation time.

@@

After a first overview of the results, GAs have quicky proved successful at optimizing the considered AM implementations. Figure \ref{fig:gas_evolution_good} shows the evolution of the score of the best individuals over generations for 10 optimizations processed for a given configuration of GAs (same operators options). 7 out of 10 optimizations reach the supposed global optimum (dashed line at the top of the plot) after 40 to 55 generations, which is not the best performance we observed later on. The performance score of the sequential approach (bottom line) has been quickly exceeded without adding any new parameter to the method (Fig. \ref{fig:gas_evolution_good}). The improvement is about 8.8\% in this case, which is substantial.

Even GAs configurations that will be considered later as inadequate, because they did not reach the supposed global optimum (Fig. \ref{fig:gas_evolution_bad}), did significantly better that the sequential approach. GAs proved promising at optimizing AMs automatically, globally, and objectively.

@@

Having identified the leading role of the mutation operator, the next sensitivity analysis focused on it. Each of the 10 different implementations (see section \ref{sec:gas}.\ref{sec:gas:operators}.\ref{sec:gas:mutation}) was tested, with different parameters for those who require some (Fig. \ref{fig:operator_mutation_score}), bringing the number of variations up to 109. Some optimizations without any mutation were also performed as a reference. Along with variants of the other operators \citep[see][for the details]{Horton2012a}, this resulted in 660 combinations (so 6,600 optimizations).

Figure \ref{fig:operator_mutation_score} show the results of this analysis and illustrates the important role of the mutation on the performance of the optimizations. Those without mutation (last box on Figure \ref{fig:operator_mutation_score}) are inferior to most mutation operators, and the scale of the influence of this operator is significantly more important than those for the other options. The details of the analysis \citep[see][]{Horton2012a} show that the other reproduction operators seem of secondary importance. This observation is in line with the work of \citet{Back1996a}, who argues for the importance of mutation over reproduction, as well as other authors (see section \ref{sec:gas}.\ref{sec:gas:operators}.\ref{sec:gas:mutation}).

@@

Then, the same optimization was performed, but for another region, sensitive to other meteorological influences (Fig. \ref{fig:operator_mutation_score_rhoneamont}), in order to assess eventual dependencies of the operator with the predictand. Even though differences can be observed with Fig. \ref{fig:operator_mutation_score_atmlevel}, it is globally the same options that perform better.

@@

The optimization does not systematically converge to the global optimum (but still often nearby), which is why it is recommended to do several optimizations in parallel in order to compare the results, analyze the convergence, and keep the best. It may be wise to consider the three mutation operators in parallel. In order to be confident in the optimized AMs, we propose using a set of the following mutation operators:

@@

It is not excluded that another global optimization method or other operators of GAs may perform still better. However, the relevance of such an approach has now been proved as it results in parameters of AMs that are relevant and fully automatically, globally and objectively established. A global optimization is the only way to take into account all the dependencies between parameters and levels of analogy.

>>>>

Analog methods (AMs) relies on the hypothesis that similar situations in terms of atmospheric circulation are likely to lead to similar local weather \citep{Lorenz1956, Lorenz1969, Duband1970, Bontron2005}. The principle consists in sampling a certain number of past situations based on different atmospheric variables (predictors) in order to build a probabilistic prediction for a local weather variable of interest (predictand). A common usage of AMs is for precipitation forecasting or downscaling \citep[eg.][]{Guilbaud1997, Bontron2005, Hamill2006, Bliefernicht2010, Marty2012, Horton2012, Radanovics2013, Chardon2014, Dayon2015, Hamill2015b, BenDaoud2016}, but AMs, or equivalent, were also used to predict temperature \citep{Radinovic1975, Woodcock1980, Kruizinga1983, DelleMonache2013, Caillouet2016}, wind \citep{Gordon1987, DelleMonache2013, DelleMonache2011, Vanvyve2015, Alessandrini2015, Junk2015, Junk2015c}, solar power \citep{Alessandrini2015a, Bessa2015}, snow avalanches \citep{Obled1980, Bolognesi1993}, insolation \citep{Bois1981}, and the trajectory of tropical cyclones \citep{Keenan1981, Sievers2000, Fraedrich2003}.

Even though the method is rather simple, it contains several parameters to optimize, such as the choice of the predictor variables, at certain pressure levels and times of the day, the spatial domains used for the comparison, as well as the analogy criteria (distance measure), and finally the number of analog situations to keep.

@@

In order to overcome these limitations, two optimization techniques were assessed. First, \citet{Horton2012a} assessed the ability of the \citet{Nelder1965a} method based on a simplex algorithm. This technique did not provide satisfying results and failed at converging toward a unique solution. The parameter space of AMs can be very complex and is inappropriate for a linear optimization technique. The conclusion was that global optimization techniques were necessary in order to calibrate most AM variants, as it is the only way to optimize all parameters of all analogy levels simultaneously. In addition, it can overcome the systematic manual assessments of all pressure levels and temporal windows. Finally, it allows assessing new degrees of freedom, such as the use of differentiated spatial windows between the pressure levels. The relevance of Genetic Algorithms (GAs) is presented here, which does not exclude that other global optimization techniques could also work. If using GAs to optimize AMs is computationally intensive, once an AM is calibrated, its use in real-time operations it very fast and lightweight.

@@

Predictors, for precipitation prediction, can be varied: e.g. geopotential heights at different pressure levels and on different temporal windows (time of observation), humidity variables, etc \citep[see][for a more detailed list]{BenDaoud2016}. The method often contains several levels of analogy, which are successive subsampling on predictors of different nature (eg. atmospheric circulation, moisture variables, vertical motion, and air temperature).

@@

	\item First level of analogy: $n_{1}$ dates are selected out of the preselected $n_{0}$, by means of an analogy ranking. The first level of analogy is often based on the atmospheric circulation when it comes to precipitation prediction. The similarity of the atmospheric circulation of the target date with the candidate situations is thus assessed on the geopotential height (at specific pressure levels, such as 500 and 1000~hPa, and over a certain spatial window, e.g. 12~h and 24~h) by means of the S1 criteria \citep[Eq. (\ref{eq:S1}), ][]{Teweles1954, Drosdowsky2003}, which is a comparison of gradients, over a certain spatial window. This criteria was found more relevant than Euclidean distances for pressure fields by different studies \citep{Wilson1980, Woodcock1980, Guilbaud1998, Bontron2004}.

@@

\begin{equation}
\label{eq:CRPSS}
CRPSS = \frac{CRPS-CRPS_{r}}{CRPS_{p}-CRPS_{r}} = 1-\frac{CRPS}{CRPS_{r}}
\end{equation}
where $CRPS_{r}$ is the CRPS value for the reference and $CRPS_{p}$ would be the one for a perfect prediction (which implies $CRPS_{p}~=~0$). An increase in CRPSS means a better prediction.

@@

A key element of GAs configuration is finding the right balance between exploration and exploitation \citep{Back1992a, Smith1997a}. Exploration is characterized by a relatively high probability to assess the regions of the parameter space that have not yet been visited. This probability must be sufficiently large at the beginning of the optimization, so that the algorithm is capable of identifying the region where the global optimum is likely located. Exploitation is characterized by a local search in an area of interest, and generally makes small movements. The latter aims at finding small improvements at the end of the optimization.

@@

GAs optimize a population of $N$ individuals. Each individual contains a chromosome, which is here a vector of the AM parameters. Genes are the individual parameters constituting the chromosome. They can be either categorical (e.g. choice of the meteorological variable), discrete (e.g. number of analogs to select), or continuous.

@@

The mutation operator modifies some gene values. It adds diversity to the population and prevents a freeze of the evolution, or a genetic drift to a local optimum. Thus, it makes the convergence to the global optimum theoretically possible \citep{Beasley1993a}, as it allows exploring beyond the current region of the parameters space by bringing new characteristics that were not present in the original population \citep{Haupt2004}.

@@

As the assessment (calculation of the objective function) of each member of the population of a generation is completely independent, it was performed in parallel on different processors of a computer. In order to perform optimizations for multiple time series, the use of a cluster is a necessity.

Some constraints need to be taken into account. For example, when a crossover or a mutation operation results in a parameter value standing out of the authorized bounds, it has to be brought back within the limits. Moreover, the parameters are of different types: continuous, discrete, or categorical (e.g. the selection of the meteorological variable). New values resulting from the optimizer need to respect the type of data it represents. Other constraints exist in between the parameters, e.g. the hours of observation of the relative humidity and the precipitable water predictors have to match when processing the moisture index.

@@

As one of our goals is to make recommendations on how to use GAs in order to optimize different AM variants, a systematic procedure was adopted. The results are summarized hereafter \citep[see][for the details]{Horton2012a}. In order to compare different configurations of GAs, the factorial design approach was applied, in a similar way as \citet{Costa2005a,Costa2007a} and \citet{Mariano2010a}. It allows isolating the effect of a parameter under different combinations of the other options. A procedure by stages was adopted, analyzing in details and in a systematic way every variant of the implemented operators, in combination with multiple other options and parameters in order to take into account eventual co-dependencies.

@@

The performances were characterized by four indicators: (i) mean performance score: average of the final scores of the 10 optimizations, (ii) convergence: the number of optimizations that converged, (iii) number of generations: characterization of the convergence speed, and (iv) number of evaluations of the objective function: characterization of the required calculation time.

@@

After a first quick assessment, GAs were found to be successful at optimizing the considered AM implementations. Figure \ref{fig:gas_evolution_good} shows the evolution of the score of the best individuals over generations for 10 optimizations processed for a given configuration of GAs (same operators options). 7 out of 10 optimizations converged to the global optimum (dashed line at the top of the plot) after 40 to 55 generations, which is not the best performance we observed later on. The performance score of the sequential approach (bottom line) has been quickly exceeded without adding any new parameter to the method (Fig. \ref{fig:gas_evolution_good}). The improvement is about 8.8\% in this case, which is substantial.

Even GAs configurations that will be considered later as inadequate, because they did not reach the global optimum (Fig. \ref{fig:gas_evolution_bad}), did significantly better that the sequential approach. GAs proved promising at optimizing AMs automatically, globally, and objectively.

@@

Having identified the leading role of the mutation operator, the next sensitivity analysis focused on it. The 10 different implementations (see section \ref{sec:gas}.\ref{sec:gas:operators}.\ref{sec:gas:mutation}) were tested, with different options (Fig. \ref{fig:operator_mutation_score}), bringing the number of variations up to 109. Some optimizations without any mutation were also performed as a reference. Along with variants of the other operators \citep[see][for the details]{Horton2012a}, this resulted in 660 combinations (so 6,600 optimizations).

Figure \ref{fig:operator_mutation_score} show the results of this analysis and illustrates the important role of the mutation on the performance of the optimizations. Those without mutation (last box on Figure \ref{fig:operator_mutation_score}) are inferior to most mutation operators, and the scale of the influence of this operator is significantly more important than those for the other options. The details of the analysis \citep[see][]{Horton2012a} show that the other reproduction operators seem of secondary importance. This observation is in line with the conclusions made by several other authors (see section \ref{sec:gas}.\ref{sec:gas:operators}.\ref{sec:gas:mutation}).

@@

% TODO: add region name and ref to map
The same experiment was performed for another region, which is sensitive to other meteorological influences (Fig. \ref{fig:operator_mutation_score_rhoneamont}), in order to assess eventual dependencies of the operator with the predictand. Even though differences can be observed with Fig. \ref{fig:operator_mutation_score_atmlevel}, it is globally the same options that perform better.

@@

The optimization does not systematically converge to the global optimum (but often close to it), which is why it is recommended to do several optimizations in parallel in order to compare the results, analyze the convergence, and keep the best. It may be wise to consider the three mutation operators in parallel. In order to be confident in the optimized AMs, we propose using a set of the following mutation operators:

@@

It is not excluded that another global optimization method or other operators of GAs may perform even better. Still, the relevance of global optimization techniques for AMs has now been proved as they result in parameters of AMs that are relevant and fully automatically, globally and objectively established. A global optimization is the only way to take into account all the dependencies between parameters and levels of analogy.
